---
title: "Glucose_counts_matrix"
author: "Renee Matthews"
date: "2025-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
			dev = c("png","pdf")
	
)
```

```{css, echo=FALSE}
pre {
  max-height: 400px;
  overflow-y: auto;
}

pre[class] {
  max-height: 200px;
}
```

```{r package loading}
library(tidyverse)
library(readr)
library(kableExtra)
library(DT)
library(readr)
library(edgeR)
library(ComplexHeatmap)
library(data.table)
library(genomation)
library(GenomicRanges)
library(readxl)
library(stringr)
library(org.Hs.eg.db)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(RColorBrewer)
library(ggpubr)
```
Data loading
```{r code for count matrix, eval=FALSE}
count_files <- list.files("C:/Users/renee/Other_projects_data/Glucose_files/counts_files", pattern = "\\.counts\\.txt$", full.names = TRUE)

counts_list <- lapply(count_files, function(file) {
   df <- read.delim(file, header = TRUE)
  
  # Rename "X" column to Geneid
  colnames(df)[1] <- "Geneid"
  
  # Rename second column to a sample name (based on file name, not bam name)
  sample_name <- tools::file_path_sans_ext(basename(file))
  colnames(df)[2] <- sample_name
  
  # Keep only Geneid and the count column
  df <- df[, c("Geneid", sample_name)]
  
  return(df)
})

merged_counts <- Reduce(function(x, y) full_join(x, y, by = "Geneid"), counts_list)
merged_counts %>%
  rename_with(~ str_remove(., "_R1\\.counts$"),
              .cols = -Geneid) %>% 
  write_tsv(.,"data/Annotation_data/Full_counts_table.tsv")
  

```
```{r functions to use}
prop_var_percent <- function(pca_result){ 
  # Ensure the input is a PCA result object
  if (!inherits(pca_result, "prcomp")) {
    stop("Input must be a result from prcomp()")
  }
  
  # Get the standard deviations from the PCA result
  sdev <- pca_result$sdev
  
  # Calculate the proportion of variance
  proportion_variance <- (sdev^2) / sum(sdev^2)*100
  
  return(proportion_variance)
}
```

```{r data loading}
sample_info_concat <- read_tsv("data/Annotation_data/organized_sample_info.tsv")
Counts_table <- read_tsv("data/Annotation_data/Full_counts_table.tsv") 

raw_counts <- Counts_table %>% 
  column_to_rownames("Geneid") %>% 
  as.matrix()
new_names <-
sample_info_concat %>% 
  mutate(new_trt= case_when(`Conc(mM)`=="0"~ "low",
                            `Conc(mM)`=="5"~ "norm",
                            `Conc(mM)`=="5+25" ~"osnorm",
                            `Conc(mM)`=="30"~ "high")) %>% 
  mutate(sex=case_when(Line=="75-1"~ "f",
                       Line=="87-1"~ "f",
                       Line=="78-1" ~"f",
                       Line=="17-3"~ "m",
                       Line=="84-1"~"m",
                       Line=="90-1"~"m")) %>% 
  mutate(new_name=paste0(Individual,"_",new_trt,"_",Time,"_",sex,"_",Replicate)) 

mapped <- select(org.Hs.eg.db,
                 keys = row.names(raw_counts),
                 columns = c("SYMBOL"),
                 keytype = "ENTREZID")

mapped_unique_symbolref <- mapped %>%
  as.data.frame() %>% 
  group_by(ENTREZID) %>%
  dplyr::slice(1) %>%
  ungroup()

```

### Initial eval of matrix

```{r hist unfilt, fig.width=10}
par(mfrow = c(1, 2))  # Set layout: 1 row, 2 columns
hist(raw_counts, main= "Raw counts across samples",
     xlab = "Raw counts",
     col=4)
hist(cpm(raw_counts, log=TRUE), 
     main = expression("Histogram of Log"[2]*" cpm unfiltered"),
     xlab = expression("Log"[2]*" counts-per-million"),
     col=4)
par(mfrow = c(1, 1))  # Reset layout to default after

boxplot(cpm(raw_counts, log=TRUE),
        main=expression("Boxplot of Log"[2]*" counts-per-million"),
                        col=4,
        names=colnames(raw_counts),
        las=2, cex.axis=.5)


```
The raw count matrix has `r length(Counts_table$Geneid)` genes in ENTREZID or NCBI_gene format.

```{r filtering low counts out}

lcpm <- cpm(raw_counts, log= TRUE)
  ### for determining the basic cutoffs
filt_raw_counts <- raw_counts[rowMeans(lcpm)> 0,]
dim(filt_raw_counts)
#tail(rownames(filt_raw_counts),n=10)
filt_raw_counts_1 <- raw_counts[rowMeans(lcpm)> 1,]
filt_raw_counts_0.5 <- raw_counts[rowMeans(lcpm)> 0.5,]
```
Filtering using row_means of the log2cpm >0 yields `r dim(filt_raw_counts)`, genes for all columns. If I had used rowMeans >0.5, I would have had `r dim(filt_raw_counts_0.5)` genes.  If I had used rowMeans >1, I would have `r dim(filt_raw_counts_1)` genes. 
```{r hist filt, fig.width=10}
par(mfrow = c(1, 2))  # Set layout: 1 row, 2 columns
hist(filt_raw_counts, main= "Raw counts across samples",
     xlab = "Raw counts",
     col=6)
hist(cpm(filt_raw_counts, log=TRUE), 
     main = expression("Histogram of Log"[2]*" cpm filtered rowMeans>0"),
     xlab = expression("Log"[2]*" counts-per-million"),
     col=6)
par(mfrow = c(1, 1))  # Reset layout to default after

boxplot(cpm(filt_raw_counts, log=TRUE),
        main=expression("Boxplot of Log"[2]*" counts-per-million of filtered set"),
                        col=6,
        names=colnames(raw_counts),
        las=2, cex.axis=.5)


par(mfrow = c(1, 2))  # Set layout: 1 row, 2 columns
hist(cpm(filt_raw_counts_0.5, log=TRUE), 
     main = expression("Histogram of Log"[2]*" cpm filtered rowMeans>0.5"),
     xlab = expression("Log"[2]*" counts-per-million"),
     col=3)
hist(cpm(filt_raw_counts, log=TRUE), 
     main = expression("Histogram of Log"[2]*" cpm filtered rowMeans >1"),
     xlab = expression("Log"[2]*" counts-per-million"),
     col=7)
par(mfrow = c(1, 1))  # Reset layout to default after

```

The distribution of log2cpm is fairly standard across all filtering steps, so I am going with the less stringent filtering of rowMeans > 0.

### Heatmap of log~2~cpm filtered genes

```{r rename columns}
## adding in renamed filt_raw_counts columns:
rename_vec <- setNames(new_names$new_name, new_names$Core_ID)
filt_raw_counts <- filt_raw_counts %>% 
  as.data.frame() %>% 
  rename_with(~ rename_vec[.x], .cols = all_of(new_names$Core_ID)) %>% 
  as.matrix()
```


```{r filtered heatmap, fig.height=14, fig.width=16}
cor_lcpm_counts_pearson <- filt_raw_counts %>% 
  cpm(., log = TRUE) %>% 
  cor(.,method = "pearson")

cor_lcpm_counts_spearman <- filt_raw_counts %>% 
  cpm(., log = TRUE) %>% 
  cor(.,method = "spearman")


anno_raw_counts <- data.frame(timeset = colnames(filt_raw_counts))

counts_corr_mat <-
anno_raw_counts %>%
  separate(timeset, into = c("indv","glucose","time","sex","rep"), sep= "_") %>% 
  mutate(indv=factor(indv, levels= c( "1","2","3","4","5","6")),
         glucose=factor(glucose, levels=c("low", "norm", "osnorm", "high")),
         time=factor(time, levels = c("2hr" ,"24hr", "1week")),
         sex=factor(sex, levels=c("f","m")),
         rep=factor(rep, levels=c("A","B")))

###pulling color palette:
cb_palette <- brewer.pal(6, "Set1") 
dr_palette <- brewer.pal(4,"Dark2")

     
 mat_colors <- list(
   indv= cb_palette,
   glucose=dr_palette,
   time=c("blue1","purple4","red4"),
   sex=c("purple", "orange"),
   rep=c("grey","grey2"))

names(mat_colors$indv)   <- unique(counts_corr_mat$indv)
names(mat_colors$time) <- unique(counts_corr_mat$time)
names(mat_colors$glucose) <- unique(counts_corr_mat$glucose)
names(mat_colors$sex) <- unique(counts_corr_mat$sex)
names(mat_colors$rep) <- unique(counts_corr_mat$rep)


htanno_full_pearson <-  ComplexHeatmap::HeatmapAnnotation(df = counts_corr_mat, col = mat_colors)
Heatmap(cor_lcpm_counts_pearson, 
        column_title="Filtered log2cpm pearson correlation",
        top_annotation = htanno_full_pearson
        )

htanno_full_spearman <-  ComplexHeatmap::HeatmapAnnotation(df = counts_corr_mat, col = mat_colors)
Heatmap(cor_lcpm_counts_spearman, 
        top_annotation = htanno_full_spearman,
        column_title="Filtered log2cpm spearman correlation")

```

### PCA analysis

```{r PCA functions prelim}
prop_var_percent <- function(pca_result){ 
  # Ensure the input is a PCA result object
  if (!inherits(pca_result, "prcomp")) {
    stop("Input must be a result from prcomp()")
  }
  
  # Get the standard deviations from the PCA result
  sdev <- pca_result$sdev
  
  # Calculate the proportion of variance
  proportion_variance <- (sdev^2) / sum(sdev^2)*100
  
  return(proportion_variance)
}

get_regr_pval <- function(mod) {
  # Returns the p-value for the Fstatistic of a linear model
  # mod: class lm
  stopifnot(class(mod) == "lm")
  fstat <- summary(mod)$fstatistic
  pval <- 1 - pf(fstat[1], fstat[2], fstat[3])
  return(pval)
}

plot_versus_pc <- function(df, pc_num, fac) {
  # df: data.frame
  # pc_num: numeric, specific PC for plotting
  # fac: column name of df for plotting against PC
  pc_char <- paste0("PC", pc_num)
  # Calculate F-statistic p-value for linear model
  pval <- get_regr_pval(lm(df[, pc_char] ~ df[, fac]))
  if (is.numeric(df[, f])) {
    ggplot(df, aes_string(x = f, y = pc_char)) + geom_point() +
      geom_smooth(method = "lm") + labs(title = sprintf("p-val: %.2f", pval))
  } else {
    ggplot(df, aes_string(x = f, y = pc_char)) + geom_boxplot() +
      labs(title = sprintf("p-val: %.2f", pval))
  }
}


```

```{r pca plots}
lcpm_filt_counts <- filt_raw_counts %>% 
  cpm(.,log = TRUE)
 
df_anno_mat <- anno_raw_counts %>%
  separate(timeset, into = c("indv","glucose","time","sex","rep"), sep= "_") %>%
  mutate(indv=factor(indv, levels= c( "1","2","3","4","5","6")),
         glucose=factor(glucose, levels=c("low", "norm", "osnorm", "high")),
         time=factor(time, levels = c("2hr" ,"24hr", "1week")),
         sex=factor(sex, levels=c("f","m")),
         rep=factor(rep, levels=c("A","B")))
pca_lcpm_filt <- prcomp(t(lcpm_filt_counts), scale. = TRUE)
pca_filt_anno <- data.frame(df_anno_mat, pca_lcpm_filt$x)
plotting_var_names <- prop_var_percent(pca_lcpm_filt)

pca_filt_anno %>% 
ggplot(.,aes(x = PC1, y = PC2, col=glucose, shape=time, group=rep))+
  geom_point(size= 3)+
   ggrepel::geom_text_repel(aes(label = interaction(indv,rep)))+
   ggtitle(expression("PCA of log"[2]*"(cpm) filtered peak set"))+
  theme_bw()+
  guides(size =4)+
  labs(y = paste0("PC 2 (",round(plotting_var_names[2],2),"%)")
       , x =paste0("PC 1 (",round(plotting_var_names[1],2),"%)"))+
  theme(plot.title=element_text(size= 14,hjust = 0.5),
        axis.title = element_text(size = 12, color = "black"))

pca_filt_anno %>% 
ggplot(.,aes(x = PC3, y = PC4, col=glucose, shape=time, group=glucose))+
  geom_point(size= 3)+
   ggrepel::geom_text_repel(aes(label = interaction(indv,rep)))+
   ggtitle(expression("PCA of log"[2]*"(cpm) filtered peak set"))+
  theme_bw()+
  guides( size =4)+
  labs(y = paste0("PC 4 (",round(plotting_var_names[4],2),"%)")
       , x =paste0("PC 3 (",round(plotting_var_names[3],2),"%)"))+
  theme(plot.title=element_text(size= 14,hjust = 0.5),
        axis.title = element_text(size = 12, color = "black"))


pca_filt_anno %>% 
ggplot(.,aes(x = PC2, y = PC3, col=glucose, shape=time, group=glucose))+
  geom_point(size= 3)+
   ggrepel::geom_text_repel(aes(label = interaction(indv,rep)))+
   ggtitle(expression("PCA of log"[2]*"(cpm) filtered peak set"))+
  theme_bw()+
  guides( size =4)+
  labs(y = paste0("PC 3 (",round(plotting_var_names[3],2),"%)")
       , x =paste0("PC 2 (",round(plotting_var_names[2],2),"%)"))+
  theme(plot.title=element_text(size= 14,hjust = 0.5),
        axis.title = element_text(size = 12, color = "black"))
```

### Contribution to variance

```{r variance by type}
facs <- c("indv", "glucose", "time", "sex")
names(facs) <- c("Individual", "Glucose level", "Time", "Sex")

for (f in facs) {
  # PC1 v PC2
  # pca_plot(pca_final_four_anno, col_var = f, shape_var = time,
  #          title = names(facs)[which(facs == f)])
  # print(last_plot())
  
  # Plot f versus PC1 and PC2
  f_v_pc1 <- gridExtra::arrangeGrob(plot_versus_pc(pca_filt_anno, 1, f))
  f_v_pc2 <- gridExtra::arrangeGrob(plot_versus_pc(pca_filt_anno, 2, f))
  f_v_pc3 <- gridExtra::arrangeGrob(plot_versus_pc(pca_filt_anno, 3, f))
  f_v_pc4 <- gridExtra::arrangeGrob(plot_versus_pc(pca_filt_anno, 4, f))
  f_v_pc5 <- gridExtra::arrangeGrob(plot_versus_pc(pca_filt_anno, 5, f))
  gridExtra::grid.arrange(f_v_pc1, f_v_pc2, f_v_pc3,f_v_pc4, f_v_pc5, ncol = 3, nrow=2, top = names(facs)[which(facs == f)])
}

```

